The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-08-14 manuscript.Rmd
revise first part of results

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016] (but see MarineGEO and PISCO), making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.

@@

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS. For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")` records) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]). For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.

<!-- this paragraph needs overhaul to make it clearer -->
For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)A). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)B). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

@@

@@

>>>>

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but see MarineGEO [@Duffy2016marinegeo] and PISCO), making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.

@@

We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively. For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). The combination of the species listed in these three sources leads to `r n_distinct(total_gom_records$worms_valid_name)` species for the GOM and `r n_distinct(total_pnw_records$worms_valid_name)` species for the PNW.

@@

Digitization of species richness is higher for the GOM compared with the PNW. A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW. While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig. \@ref(fig:database-coverage)A), `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig. \@ref(fig:database-coverage)B). Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_gom_records$worms_valid_name)`; PNW:
`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_pnw_records$worms_valid_name)`).

@@

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.

@@

<!-- Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases. -->

<!-- First, we compared the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases. we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS. For the Gulf of Mexico, we retrieved -->
